\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-09-19 demystify.tex
Updates from Overleaf

International students not familiar with US universities often put them into \emph{two} bins:  (i) very top schools that they dream about such as Stanford, MIT, Princeton, Harvard and (ii) everything else.  In many cases, they use resources such as US News, but their ranking is \emph{highly} questionable~\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes these students evaluate CS programs using the reputations non-CS programs such as medical, math, or physics.
In some cases they rank universities based on popular states they know in the US, e.g., California and New York.  Clearly, there are so many thing wrong with these methods.

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about has no one works in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured). They might also expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research and more involved with administrative duties or with their startup companies.

You can learn about faculty and their level of research activity through the faculty's website and CSRankings.
Ultimately, choose one that fits you the most by communicating with them, meeting them, asking them questions, even talking to their students.

@@

If you want to contact a prof. to ask about \emph{research opportunities}, or \emph{GTA/GRA} support, then \emph{yes}, I believe you should---it is \emph{worth it}. However, you need to do it in a right way.

@@

    The first 2 years you spend on coursework (somewhat equivalent to an MS study), finding an adviser, learning research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. 
    In many cases you might take a summer or two off to do internship to get additional research opportunities.
    The PhDComics figure on the right shows the ``ambition'' level of a PhD student over their years of study (they miss the 6-7th Year where the ambition is \emph{``Just let me graduate''}).

@@

\subsection{How much do \emph{you} cost?}\label{sec:ra-cost}
PhD students often wonder why their salary is so low compared to ludicrous grants their advisors get or why their offer letters sometime mentioned that their benefits worth way more than what they actually receive (i.e., stipend).  This section aims to shed some light to these questions.

@@

\paragraph{Acknowledgement} Many people have contributed to this document.
Profs. Craig Yu (GMU), Hakan Aydin (GMU), and Hung Le (UMass) provided valuable input in the early version. Other GMU faculty members also have provided feedback and contributions.  Many students including Didier (GMU), Thanh (Melbourne), and Dat (Melbourne) also have contributed valuable questions and feedback. 
\textbf{Thank you!}

>>>>

International students not familiar with US universities often put them into \emph{two} bins:  (i) very top schools that they dream about such as Stanford, MIT, Princeton, Harvard and (ii) everything else.  In many cases, they use resources such as rankings from US News, which are \emph{highly} questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes these students evaluate CS programs using the reputations non-CS programs such as medical, math, or physics.
They even rank universities based on popular states they know in the US, e.g., California and New York.  Clearly, there are so many thing wrong with these methods.

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenure). They might also expect you to figure things out yourself, i.e., so you need to be very independent.  Some tenured faculty are also no longer active in research and more involved with administrative duties or with their startup companies (this means they will likely not take new students).

You can learn about faculty and their level of research activity through the faculty's website or even from CSRankings.
Ultimately, choose one that fits you the most by communicating with them, meeting them, asking them questions, and even talking to their students.

@@

If you want to contact a prof. to ask about \emph{research opportunities}, or \emph{GTA/GRA} support, then \emph{yes}, I believe you should---it is \emph{worth it}. However, you need to do it the \emph{right way}.

@@

    The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, learn how to do research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. 
    In many cases you might take a summer or two off to do internship to get additional research opportunities.
    The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6-7th year where the ambition is \emph{``Just let me graduate''}).

@@

\subsection{How much do \emph{you} cost?}\label{sec:ra-cost}
PhD students often ask why their salary is so low compared to ludicrous grants their advisors get. They also wonder why their offer letters sometime mentioned that their benefits worth way more than what they actually receive (i.e., stipend).  This section aims to shed some light to these questions.

@@

\paragraph{Acknowledgement} Many people have contributed to this document.
Profs. Craig Yu (GMU), Hakan Aydin (GMU), and Hung Le (UMass) provided valuable input in the early version. Other GMU faculty members also have provided feedback and contributions.  Many students including Didier (GMU), Thanh (Melbourne), and Dat (Melbourne) have contributed valuable questions and feedback. 
\textbf{Thank you!}
